\textbf{Local Search Algorithms for the \\ Maximum Carpool Matching Problem} \thanks{A preliminary version was presented at the 25th Annual European Symposium on Algorithms, 2017}

### 2019-12-06 response.tex

\documentclass[11pt]{article}

\usepackage{amsmath,latexsym,amssymb} 
\usepackage{fullpage,ifthen}
\usepackage{color,soul,colordvi}
\usepackage{mdwlist,paralist}

%\renewcommand{\baselinestretch}{1.02}

%\parskip=\smallskipamount

\sloppy

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newcommand{\abs}[1]{\left| #1 \right|}
\newcommand{\ceil}[1]{\left\lceil {#1} \right\rceil}
\newcommand{\floor}[1]{\left\lfloor {#1} \right\rfloor}
\newcommand{\paren}[1]{\left( #1 \right)}
\newcommand{\set}[1]{\left\{ #1 \right\}}

\newcommand{\eps}{\varepsilon}

\newcommand{\calC}{\mathcal{C}}
\newcommand{\sigmad}{\sigma_{\$}}

\newcommand{\ans}[1]{\begin{quotation}{\noindent \sl {#1}}\end{quotation}}

\newcommand{\dror}[1]{\sethlcolor{yellow}\hl{Dror: #1}}
\newcommand{\ben}[1]{\sethlcolor{green}\hl{Ben: #1}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\title{\textbf{Local Search Algorithms for the \\ Maximum Carpool Matching
  Problem} \\
  {\Large Response to Reviewers: Submission ALGO-D-00287}}

\author{Gilad Kutiel \and Dror Rawitz}

\maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

We thank the referee for the time that was spent on reviewing our
paper and for their thoughtful comments.  The paper was revised
according to these comments.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\medskip

\section*{Response to Review}

\begin{enumerate}

\item \textbf{Comment}: There are multiple graphs used throughout the
  paper: a directed input graph, the related undirected graph
  obtained by ignoring edge orientations, and a graph $H$ that relates
  stars to one another. It would help the presentation if it were made
  completely clear at each stage which graph is being discussed.

\item[] \textbf{Response}: We now use calligraphic letters in the
  definition of the star graph, namely the star graph is denoted by
  $\mathcal{H} = (\mathcal{S}(M^*),\mathcal{E})$.

\item \textbf{Comment}: In the proof of Lemma 1, I do not see how the
  condition that $M_S \cup M_T = M(S \cup T) \cup M(S \cap T)$ is
  sufficient to finish the proof when $M(S \cup T)$ and $M(S \cap T)$
  are not disjoint. In that case the weights of elements in the
  intersection will be counted only once. The later proof seems to
  actually account for this.

\item[] \textbf{Response}: We now use $M(S \cup T) \uplus M(S \cap
  T)$.

\item \textbf{Comment}: In the proof of Lemma 8, what is meant by an
  edge of $M$ ”intersecting” $V(\mathcal{T})$?  Presumably you mean an
  edge of $M$ that is incident on one of the vertices of
  $V(\mathcal{T})$?

\item[] \textbf{Response}: We added a sentence about this issue in the
  beginning of the proof of Lemma~8.

\item \textbf{Comment}: It would be good to emphasize that the set of
  stars induced by $M^*$ includes also all of the single-vertex
  ``stars -- corresponding to $Z_M$.  This was not so clear to me on
  the first reading.

\item[] \textbf{Response}: This is now emphasized in the paragraph
  that defines $\mathcal{S}(M)$.

\item \textbf{Comment}: In the proof of Lemma~9, please point out the
  usage of Lemma~6, which bounds the degree of $H$, when you use Lemma~7.

\item[] \textbf{Response}: Done.

\item \textbf{Comment}: In the proof of Lemma~9, you write ”Since each
  such vertex set is connected and contains at most
  $\frac{k}{c_{\max}}$ stars, it follows that $|A(\mathcal{X})| \leq
  k$.''  However, I do not see how the vertex set being connected is
  relevant for $|A(\mathcal{X})|$. I agree that it is used to apply
  Lemma~8 later.

\item[] \textbf{Response}: Corrected.

\item \textbf{Comment}: In the tight example (Figure~7) it is stated
  that the solution can be improved by local search only if removes
  all of the edges.  This is not true; please see the last page of
  this document for an improvement that does not remove every edge.
  Probably one needs to rearrange the edges connecting leaves
  appropriately.  The example given by Hurkens and Schrijver%
%
  \footnote{``On the size of systems of sets every $t$ of which have an
    SDR, with an application to the worst-case ratio of heuristics for
    packing problems'', SIAM J. Discrete Math., 2(1), 6872.}
%
  might be useable here instead.  In any case, please give some
  details for a general construction or at least provide details of
  how the example can be generalized.

\item[] \textbf{Response}: We removed this figure from the paper.

\item \textbf{Comment}: Details should be given for the proofs of
  Theorems~5, 6, and~7.  Since the group variant already generalizes
  the standard setting why not just present proofs for the general
  case?

\item[] \textbf{Response}: We moved the material from Section 4 to
  sections 2.1, 2.2 and 3.2, respectively.

\end{enumerate}

%%%%%

\subsection*{Minor comments}

\begin{enumerate}

\item \textbf{Comment}: Page 3: The best known approximation for
  weighted $k$-set packing is in fact $\frac{k+1}{2}$, due to Piotr Berman.%
%
\footnote{``A $d/2$ Approximation for Maximum Weight Independent Set
  in $d$-Claw Free Graphs'' in SWAT 2000}

\item[] \textbf{Response}: A reference was added, and the last
  sentence of the paragraph was corrected accordingly.

\item \textbf{Comment}: Page 5, line 56: $v$ is missing from the
  definition of $N^{\text{in}}$ and $N^{\text{out}}$.

\item[] \textbf{Response}: Corrected.

\item \textbf{Comment}: Page 6, line 33. Again, $v$ is missing from
  $N^{\text{in}}$ and $N^{\text{out}}$.

\item[] \textbf{Response}: Corrected.

\item \textbf{Comment}: Page 7, caption of Figure 3: it seems that the
  last two edges listed in $\Gamma(2,3)$ are incorrect.

\item[] \textbf{Response}: The example was correct.  We corrected the
  definition of $\Gamma(u,v)$.

\item \textbf{Comment}: Page 8, Figure 5: are the double arrows in
  this figure intentional? If so, I do not understand what is meant by
  selecting ``all arcs of length 1.''

\item[] \textbf{Response}: The figure and caption were corrected.

\item \textbf{Comment}: Page 8, line 34: $N_{\overline{S}}$ is used
  here instead of the previous notation $N^{\text{in}}_S(v)$.  Also,
  it might be worthwhile to point out at the end that the vertices
  will not be charged again \emph{because the stars are disjoint in
    $M^*$}.

\item[] \textbf{Response}: Corrected.

\item \textbf{Comment}: Page 12, line 28: $\deg_M(u)$ should be
  $\deg_M(v)$.

\item[] \textbf{Response}: Corrected.

\item \textbf{Comment}: Page 12, line 39-41: Starting with ``Observe
  that if an edge \ldots,'' I do not follow why this condition is
  helpful or necessary for $M'$ to be a valid carpool matching.

\item[] \textbf{Response}: It helps to explain the transition from
  $\deg_M(V(\mathcal{T}))$ to $|M \cap A(\mathcal{T})|$.

\item \textbf{Comment}: Page 13, line 30: at the end of the proof it
  would be good remind that the sets $A(\mathcal{T})$ considered form
  a partition of all the edges (and so also of the edges in both $M$
  and $M^*$, and summing over them completes the proof.

\item[] \textbf{Response}: Done.

\end{enumerate}

\end{document}

### 2019-12-06 carpool.tex
adding astrix near my name

%\renewcommand{\floatpagefraction}{0.8}

@@

\newtheorem{lemma}{Lemma}
\newtheorem{observation}[lemma]{Observation}
\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}
\newtheorem{property}{Property}
\newtheorem{corollary}{Corollary}
\newtheorem{proposition}{Proposition}

@@

@@

\author{%
Gilad Kutiel%
\thanks{Department of Computer Science, Technion, Haifa 32000, Israel.} \\
\texttt{gkutiel@cs.technion.ac.il}
\and
Dror Rawitz%
\thanks{Faculty of Engineering, Bar Ilan University, Ramat Gan 52900, Israel.
Supported in part by the Israel Science Foundation (grant no.~497/14).} \\
\texttt{dror.rawitz@biu.ac.il}
}

@@

\end{tikzpicture}
\caption{An instance containing a directed graph with capacities on
  the vertices and weights on the arcs. }
\end{subfigure}
%
\hfill
%
\begin{subfigure}[t]{.5\linewidth}
\centering
\begin{tikzpicture}[every node/.style={default node}, ->, very thick]

@@

\end{tikzpicture}
\caption{A feasible carpool matching with total weight
of 23.  $P_M$ is the set of blue vertices, and $D_M$ is the set of
red, dashed vertices, and $Z_M$ contains only the dotted, black
vertex.}
\end{subfigure}
\caption[]{
\label{fig:carpool}
A \carpool example.
%(\subref{carpool:input}) an instance containing a
%directed graph with capacities on the vertices and weights on the
%arcs.  (\subref{carpool:output}) a feasible matching with total weight
%of 23.  $P_M$ is the set of blue vertices, and $D_M$ is the set of
%red, dashed vertices, and $Z_M$ contains only the dotted, black
%vertex.
}
\end{figure}

@@

Another related problem is \textsc{$k$-Set Packing}, where one is
given a collection of weighted sets, each containing at most $k$
elements, and the goal is to find a maximum weight subcollection of
disjoint sets.  Chandra and Halld\'orsson~\cite{chandra2001greedy}
presented a $\frac{3}{2(k+1)}$-approximation algorithm for this
problem.
%
\carpool can be seen as a special case of \textsc{$k$-Set Packing}
with $k = \cmax + 1$.  Consider a subset of vertice $U$ of size at
most $k$.  Observe that each subset of vertices has an optimal
internal assignment of passenger to drivers.  Let the weight of this
assignment be the profit of $U$, denote by $p(U)$.  If $k = O(1)$,
$p(U)$ can be computed for every $U$ of size at most $k$ in polynomial
time.  The outcome is a \textsc{$k$-Set Packing} instance.  This leads
to a $\frac{3}{2(\cmax+2)}$-approximation algorithm
%for the case where
when $\cmax = O(1)$.

@@

\paragraph*{Our contribution.}
%
Section~\ref{sec:approx} contains approximation algorithms
for \carpool.  First, in Section~\ref{sec:sub} we show that \carpool
can be formulated as an unconstrained submodular maximization problem,
thus it has a $\frac{1}{2}$-approximation algorithm due
to~\cite{BFNS15,buchbinder2016deterministic}.
%
We present a local search algorithm for \carpool which repeatedly
checks whether the current carpool matching can be improved by means
of a star centered at a vertex, and it terminates when such a step is
not possible.
%
The approximation ratio of this algorithm is $\half$ if weights are
polynomially bounded, and its ratio is $\half-\eps$ in general.

In Section~\ref{sec:cmax} we consider \carpool with bounded maximum
capacity.   
%
In Section~\ref{sec:hardness} we show that \carpool is APX-hard even
for undirected and unweighted instances with $\Delta \leq b$, for any
$b \geq 3$.
%
In Section~\ref{sec:local} we provide another local search algorithm,
whose approximation ratio is $\half + \inv{2\cmax} - \eps$, for any
$\eps>0$, for unweighted \carpool, where $\cmax \eqdf \max_{v \in V}
c(v)$.  Given a parameter $k$, our algorithm starts with the empty
carpool matching.  Then, it repeatedly tries to find a better matching
by replacing $t \leq k$ arcs in the current solution by $t+1$ arcs
that are not in the solution.
%
We show that our analysis is tight.
%
We also note that our algorithm falls within the local search family
defined in~\cite{ACKK09}.  However, on undirected and uncapaciated
instances we have that $\cmax = \Delta$, and as mentioned above the
result from~\cite{ACKK09} does not hold in bounded degree graphs.

Finally, Section~\ref{sec:group} discusses \gcp.  We show that the
unconstrained submodular maximization formulation for \carpool does
not work for \gcp.  We show, however, that this problem still admits a
$(\frac{1}{2} -\varepsilon)$-approximation algorithm by extending our
first local search algorithm.  In addition, we show that the second
local search algorithm generalizes to unweighted \gcp with the same
approximation ratio.

@@

We present two algorithms for \carpool: a $\frac{1}{2}$-approximation
algorithm that is based on formulating the problem as an unconstrained
submodular maximization problem and a local search $(\frac{1}{2} -
\varepsilon)$-approximation algorithm.  While the latter does not
improve upon the former, it will be shown (in Section~\ref{sec:group})
that it can be generalized to \gcp without decreasing the
approximation ratio.

@@

Consider the function $\barw: 2^V \to \R$, where $\barw(S) \eqdf
w(M(S)) = \sum_{e \in M(S)} w(e)$.  Observe that $\barw(\emptyset)
= \barw(V) = 0$, and that $\barw$ is not monotone.
%
In the next lemma we prove that $\bar{w}$ is a \emph{submodular set
function}.  Recall that a function $f$ is submodular if $f(S) +
f(T) \geq f(S \cup T) + f(S \cap T)$ for every two sets $S$ and $T$ in
the domain of $f$.

\begin{lemma}
$\barw$ is submodular.
\end{lemma}
\begin{proof}
Consider any two subsets $S, T \subseteq V$.  We show that $\barw(S)
+ \barw(T) \geq \barw(S \cup T) + \barw(S \cap T)$.
%
Let $M(S \cup T)$ and $M(S \cap T)$ be optimal carpool matchings with
respect to $S \cup T$ and $S \cap T$.
%
To prove the lemma we construct two feasible carpool matchings $M_S$
and $M_T$ such that $M_S \subseteq (V \setminus S) \times S$,
$M_T \subseteq (V \setminus T) \times T$, and $M_S \cup M_T = M(S \cup
T) \cup M(S \cap T)$.
%
The lemma follows, since $\barw(S) \geq w(M_S)$ and $\barw(T) \geq
w(M_T)$.

First, add all the edges in $M(S \cup T)$ entering $S \setminus T$ to
$M_S$.  Similarly, add all the edges in $M(S \cup T)$ entering
$T \setminus S$ to $M_T$.  Observe that $\din[M_S](v) = \din[M(S \cup
T)](v) \leq c(v)$, for every $v \in S \setminus T$ and that
$\din[M_T](v) = \din[M(S \cup T)](v) \leq c(v)$, for every $v \in
T \setminus S$.
%
Next, add the edges in $M(S \cap T)$ leaving $T \setminus S$ to $S$
and add the edges in $M(S \cap T)$ leaving $S \setminus T$ to $T$.
%
It remains to distribute the edges leaving $V \setminus (S \cup T)$
and entering $S \cap T$ in both $M(S \cup T)$ and $M(S \cap T)$.  Note
that there may exist edges $(v,u)$, where $v \not\in S \cup T$, and
$u \in S \cap T$ such that $(v,u) \in M(S \cup T)$ and $M(S \cap T)$.
We refer to this edges as \emph{duplicate} edges.
%
We add all edges leaving $V \setminus (S \cup T)$ and entering $S \cap
T$ in $M(S \cap T)$ to $M_S$.  Notice that this is possible, since
after this addition we have that $\din[M_S](v) \leq \din[M(S \cap
T)](v) \leq c(v)$, for every vertex $v \in S \cap T$.
%
Then we add all duplicate edges in $M(S \cup T)$ to $M_T$.
%
The remaining edges are distributed between $M_S$ and $M_T$ without
violating capacities.  This can be done, since $\din[M(S \cup T)](v)
+ \din[M(S \cap T)](v) \leq 2c(v)$, for every $v \in S \cap T$.
%
Figure~\ref{fig:sub} contains a graphical representation of the above
edge distribution.
\end{proof}

@@

@@

In this section we give a local search $(\half-\eps)$-approximation
algorithm for \carpool.  This algorithm repeatedly checks whether the
current carpool matching $M$ can be improved by means of a star
centered at a vertex $v$.  The profit from this star is the total
weight of the arcs in the star, and the cost is the total weight of
lost arcs (e.g., arcs from passengers to drivers that became
passengers of $v$).  If the profit is larger than the cost, then an
improvement step is performed.  The algorithm terminates when such a
step is not possible.
%
We remind the reader that this algorithm will be extended to \gcp in
Section~\ref{sec:group}.

We need a few definitions before presenting our algorithm.
%
Given a directed graphs $G = (V,A)$, define $\nin[] \eqdf \set{u :
(u,v) \in A}$ and $\nout[] \eqdf \set{u : (v,u) \in A}$.
%
Let $M$ be a feasible carpool matching.  The weight $w_M(v)$ of a
vertex $v$ with respect to $M$ is the sum of the weights of the arcs
in $M$ that are incident on $v$, namely
\[
\textstyle
w_M(v)
\eqdf w(M \cap \nin[]) + w(M \cap \nout[])
=     \sum_{(u, v) \in M} w(u, v) + \sum_{(v, u) \in M} w(v, u)
~.
\]
For a subset of vertices $U \subseteq V$ we define
$w_M(U) \eqdf \sum_{v \in U} w_M(v)$.

@@

\begin{observation}
\label{lm:val-twice}
$w_M(V) = 2 w(M)$.
%\sum_{v \in V} w_M(v) = 2 \sum_{e \in M} w(e)$.
\end{observation}
\begin{proof}
\(
\displaystyle
\sum_{v \in V} w_M(v)
= \sum_{v \in V} \sum_{(u, v) \in M} w(u, v) +
    \sum_{v \in V} \sum_{(v, u) \in M} w(v, u) 
= 2 \sum_{e \in M} w(e)
\).
\end{proof}

Denote by $\delta(u,v)$ the difference between the weight of the arc
and the weight of its source vertex, that is:
\[
\delta_M(u, v) \eqdf w(u, v) - w_M(u)
~.
\]
For a subset $S \subseteq A$ of arcs define
$\delta(S) \eqdf \sum_{(u,v) \in S} \delta(u,v)$.

@@

\begin{definition}
A subset $S_v \subseteq A \cap (V \times \{v\})$ is
an \emph{improvement} with respect to a carpool matching $M$, if%
%the following conditions hold:%
\begin{inparaenum}[(i)]
\item $\abs{S_v} \leq c(v)$, and
\item $\delta_M(S_v) > w_M(v)$.
\end{inparaenum}
Furthermore, if there exists an improvement for a vertex $v$, we say
that vertex $v$ can be \emph{improved}.
\end{definition}

Given an arc $(u,v) \in A$, let $\inc(u,v)$ be the set of arcs that
incident $(u, v)$, namely define
\(
\inc(u,v) \eqdf (\nin[](u) \times \{u\}) \cup (\{v\} \times \nout[](v))
%~.
\).
If $S$ is a set of arcs, then $\inc(S) \eqdf \bigcup_{(u,v) \in
S} \inc(u,v)$.
%
Figure~\ref{fig:defs} depicts all the above definitions.

@@

We are now ready to describe our local search algorithm, which is
called \textbf{StarImprove} (Algorithm~\ref{alg:grd}).  It starts with
an empty carpool matching $M$, and in every iteration it looks for a
vertex that can be improved.  If there exists such a vertex $v$, then
the algorithm removes the arcs that are incident on it from $M$, and
adds the arcs in $S_v$.  The algorithm terminates when no vertex can
be improved.  Figure~\ref{fig:improvement} depicts an improvement
step.

@@

\end{tikzpicture}
\caption{The matching after improving vertex 3.}
\end{subfigure}
\caption[]{An improvement example.}
\label{fig:improvement}
\end{figure}

@@

Consider a driver $v \in D_{M^*}$, and let $S = (V \times \{v\}) \cap
M^*$.  By lemma~\ref{lm:no improve} we know that $w(S) \leq w_M(v) +
w_M(\nin[S](v))$, thus we can pay for $S$, using the money on $v$ and
on $N^-_S(v)$.  Clearly, these vertices will not be charged again.
\end{proof}

We show that our analysis is tight using in Figure~\ref{fig:grd
worst}.

\begin{figure}[t]
\centering
\begin{tikzpicture}[every node/.style={default node}, <->]

@@

\end{tikzpicture}
\caption[]{%
Consider a path with $2n + 1$ arcs, and alternating arc weights (2 and
1), if \textbf{StarImprove} selects all arcs of weight 1, then no
further improvement can be done and the value of the matching is $n +
1$, while the optimal matching has value of $2n$.}
\label{fig:grd worst}
\end{figure}

@@

\section{Group Carpool}
\label{sec:group}

>>>>

\renewcommand{\floatpagefraction}{0.8}

@@

\newtheorem{lemma}{Lemma}
\newtheorem{observation}[lemma]{Observation}
\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}
\newtheorem{property}{Property}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}{Proposition}

@@

\newcommand{\calH}{\mathcal{H}}
\newcommand{\calE}{\mathcal{E}}

@@

\author{%
Gilad Kutiel\textsuperscript{*}%
\thanks{Department of Computer Science, Technion, Haifa 32000, Israel.} \\
\texttt{gkutiel@cs.technion.ac.il}\textsuperscript{*}
\and
Dror Rawitz%
\thanks{Faculty of Engineering, Bar Ilan University, Ramat Gan 52900, Israel.
Supported in part by the Israel Science Foundation (grant no.~497/14).} \\
\texttt{dror.rawitz@biu.ac.il}
}

@@

\end{tikzpicture}
\caption{An instance containing a directed graph with capacities on
  the vertices and weights on the arcs. }
\end{subfigure}
%
\hfill
%
\begin{subfigure}[t]{.45\linewidth}
\centering
\begin{tikzpicture}[every node/.style={default node}, ->, very thick]

@@

\end{tikzpicture}
\caption{A feasible carpool matching with total weight of 23.  $P_M$
  is the set of blue, solid vertices, and $D_M$ is the set of red,
  dashed vertices, and $Z_M$ contains only the dotted, black vertex.}
\end{subfigure}
\caption[]{
\label{fig:carpool}
A \carpool example.
%(\subref{carpool:input}) an instance containing a
%directed graph with capacities on the vertices and weights on the
%arcs.  (\subref{carpool:output}) a feasible matching with total weight
%of 23.  $P_M$ is the set of blue vertices, and $D_M$ is the set of
%red, dashed vertices, and $Z_M$ contains only the dotted, black
%vertex.
}
\end{figure}

@@

Another related problem is \textsc{$k$-Set Packing}, where one is
given a collection of weighted sets, each containing at most $k$
elements, and the goal is to find a maximum weight subcollection of
disjoint sets.  Chandra and Halld\'orsson~\cite{chandra2001greedy}
presented a $\frac{3}{2(k+1)}$-approximation algorithm for this
problem, and Berman~\cite{Berman00} gave a
$\frac{2}{k+1}$-approximation algorithm.
%
\carpool can be seen as a special case of \textsc{$k$-Set Packing}
with $k = \cmax + 1$.  Consider a subset of vertice $U$ of size at
most $k$.  Observe that each subset of vertices has an optimal
internal assignment of passenger to drivers.  Let the weight of this
assignment be the profit of $U$, denote by $p(U)$.  If $k = O(1)$,
$p(U)$ can be computed for every $U$ of size at most $k$ in polynomial
time.  The outcome is a \textsc{$k$-Set Packing} instance.  This leads
to a $\frac{2}{\cmax+2}$-approximation algorithm for \carpool for the
case where $\cmax = O(1)$.

@@

\paragraph*{Our contribution.}
%
Section~\ref{sec:approx} contains approximation algorithms for
\carpool.  First, in Section~\ref{sec:sub} we show that \carpool can
be formulated as an unconstrained submodular maximization problem,
thus it has a $\frac{1}{2}$-approximation algorithm due
to~\cite{BFNS15,buchbinder2016deterministic}.  On the other hand, we
show that the unconstrained submodular maximization formulation for
\carpool does not work for \gcp.
%
We present a local search algorithm for \carpool which repeatedly
checks whether the current carpool matching can be improved by means
of a star centered at a vertex, and it terminates when such a step is
not possible.
%
The approximation ratio of this algorithm is $\half$ if weights are
polynomially bounded, and its ratio is $\half-\eps$ in general.  This
algorithm extends to \gcp with the same approximation ratio.
%We show, however, that this problem still
%admits a $(\frac{1}{2} -\varepsilon)$-approximation algorithm by
%extending our first local search algorithm.

In Section~\ref{sec:cmax} we consider \carpool with bounded maximum
capacity.   
%
In Section~\ref{sec:hardness} we show that \carpool is APX-hard even
for undirected and unweighted instances with $\Delta \leq b$, for any
$b \geq 3$.
%
In Section~\ref{sec:local} we provide another local search algorithm,
whose approximation ratio is $\half + \inv{2\cmax} - \eps$, for any
$\eps>0$, for unweighted \carpool, where $\cmax \eqdf \max_{v \in V}
c(v)$.  Given a parameter $k$, our algorithm starts with the empty
carpool matching.  Then, it repeatedly tries to find a better matching
by replacing $t \leq k$ arcs in the current solution by $t+1$ arcs
that are not in the solution.
%
%We show that our analysis is tight.
%
We also note that our algorithm falls within the local search family
defined in~\cite{ACKK09}.  However, on undirected and uncapaciated
instances we have that $\cmax = \Delta$, and as mentioned above the
result from~\cite{ACKK09} does not hold in bounded degree graphs.
%
In addition, we show that our local search algorithm generalizes to
unweighted \gcp with the same approximation ratio.

@@

We present two algorithms for \carpool: a $\frac{1}{2}$-approximation
algorithm that is based on formulating the problem as an unconstrained
submodular maximization problem and a local search $(\frac{1}{2} -
\varepsilon)$-approximation algorithm.  While the latter does not
improve upon the former, it will be shown
%(in Section~\ref{sec:group})
that it can be generalized to \gcp without decreasing the
approximation ratio.

@@

Consider the function $\barw: 2^V \to \R$, where
\[
\barw(S) \eqdf w(M(S)) = \sum_{e \in M(S)} w(e)
~.
\]
Observe that $\barw(\emptyset) = \barw(V) = 0$, and that $\barw$ is
not monotone.
%
In the next lemma we prove that $\bar{w}$ is a \emph{submodular set
function}.  Recall that a function $f$ is submodular if $f(S) +
f(T) \geq f(S \cup T) + f(S \cap T)$ for every two sets $S$ and $T$ in
the domain of $f$.

\begin{lemma}
$\barw$ is submodular.
\end{lemma}
\begin{proof}
Consider any two subsets $S, T \subseteq V$.  We show that $\barw(S)
+ \barw(T) \geq \barw(S \cup T) + \barw(S \cap T)$.
%
Let $M(S \cup T)$ and $M(S \cap T)$ be optimal carpool matchings with
respect to $S \cup T$ and $S \cap T$.
%
To prove the lemma we construct two feasible carpool matchings $M_S$
and $M_T$ such that $M_S \subseteq (V \setminus S) \times S$, $M_T
\subseteq (V \setminus T) \times T$, and $M_S \cup M_T = M(S \cup T)
\biguplus M(S \cap T)$, where the RHS can be a multiset.
%
The lemma follows, since $\barw(S) \geq w(M_S)$ and $\barw(T) \geq
w(M_T)$.

First, add all the edges in $M(S \cup T)$ entering $S \setminus T$ to
$M_S$.  Similarly, add all the edges in $M(S \cup T)$ entering
$T \setminus S$ to $M_T$.  Observe that $\din[M_S](v) = \din[M(S \cup
T)](v) \leq c(v)$, for every $v \in S \setminus T$ and that
$\din[M_T](v) = \din[M(S \cup T)](v) \leq c(v)$, for every $v \in
T \setminus S$.
%
Next, add the edges in $M(S \cap T)$ leaving $T \setminus S$ to $S$
and add the edges in $M(S \cap T)$ leaving $S \setminus T$ to $T$.
%
It remains to distribute the edges leaving $V \setminus (S \cup T)$
and entering $S \cap T$ in both $M(S \cup T)$ and $M(S \cap T)$.  Note
that there may exist edges $(v,u)$, where $v \not\in S \cup T$, and $u
\in S \cap T$ such that $(v,u) \in M(S \cup T)$ and $(v,u) \in M(S
\cap T)$.  We refer to this edges as \emph{duplicate} edges.
%
We add all edges leaving $V \setminus (S \cup T)$ and entering $S \cap
T$ in $M(S \cap T)$ to $M_S$.  Notice that this is possible, since
after this addition we have that $\din[M_S](v) \leq \din[M(S \cap
T)](v) \leq c(v)$, for every vertex $v \in S \cap T$.
%
Then we add all duplicate edges in $M(S \cup T)$ to $M_T$.
%
The remaining edges are distributed between $M_S$ and $M_T$ without
violating capacities.  This can be done, since $\din[M(S \cup T)](v)
+ \din[M(S \cap T)](v) \leq 2c(v)$, for every $v \in S \cap T$.
%
Figure~\ref{fig:sub} contains a graphical representation of the above
edge distribution.
\end{proof}

@@

Now consider \gcp.  Recall that in this variant of \carpool, in
addition to the capacity function $c:V \to \N$, and the weight
function $w:A \to \R^+$, we are also given a size function $s:V \to
\N$, and the constraint $\din[M](v) \leq c(v)$ is replaced with the
constraint $\sum_{u:(u,v) \in M} s(u) \leq c(v)$.
%
%The motivation for this variant of the problem is when we want to
%allow a group of people to travel together in the same car, thus $s$
%is the size of the group.
%
We show that this variant of the problem does not fit the submodular
maximization formulation as defined for \carpool.
%
Recall the submodular maximization formulation given above,
%in Section~\ref{sec:sub},
namely $\barw: 2^V \to \R$, where $\barw(S)
\eqdf w(M(S))$ and $M(S)$ is the maximum weight carpool matching that
satisfies $D_{M(S)} \subseteq S \subseteq V \setminus P_{M(S)}$.
%
Figure~\ref{fig:not submodular} contains an instance that shows that
the function $\barw$ is not submodular anymore.

\begin{figure}[!t]
\begin{center}
\begin{tikzpicture}[every node/.style={default node},thick]

\node(1) at (-2,0){2};
\node(2) at (0,0){1};
\node(3) at (2,0){3};
\node(4) at (0,2){4};

\draw[->] (1) -- (2);
\draw[->] (3) -- (2);
\draw[->] (4) -- (2);

\draw[dashed,red] 
(2.west) to[out=90,in=90] node[label above] {B}
(3.east) to[out=-90, in=-90]
(2.west)
;
\end{tikzpicture}
\end{center}
\caption{An unweighted \carpool instance: $c(1) = 2$, $s(2) = s(3) =
  1$, and $s(4) = 2$.  We have that $\barw(A) + \barw(B) = 2 < 3 =
  \barw(A \cup B) + \barw(A \cap B)$.}
\label{fig:not submodular}
\end{figure}

@@

In this section we give a local search $(\half-\eps)$-approximation
algorithm for \carpool.  This algorithm repeatedly checks whether the
current carpool matching $M$ can be improved by means of a star
centered at a vertex $v$.  The profit from this star is the total
weight of the arcs in the star, and the cost is the total weight of
lost arcs (e.g., arcs from passengers to drivers that became
passengers of $v$).  If the profit is larger than the cost, then an
improvement step is performed.  The algorithm terminates when such a
step is not possible.
%
At the end of the section we show that this algorithm extends to \gcp.

%We remind the reader that this algorithm will be extended to \gcp in
%Section~\ref{sec:group}.

We need a few definitions before presenting our algorithm.
%
Given a directed graph $G = (V,A)$, define $\nin[](v) \eqdf \set{u :
  (u,v) \in A}$ and $\nout[](v) \eqdf \set{u : (v,u) \in A}$.
%
Given a feasible carpool matching $M$, the weight $w_M(v)$ of a vertex
$v$ with respect to $M$ is the sum of the weights of the arcs in $M$
that are incident on $v$, namely
\[
w_M(v)
\eqdf w(M \cap \nin[](v)) + w(M \cap \nout[](v))
=     \sum_{(u, v) \in M} w(u, v) + \sum_{(v, u) \in M} w(v, u)
~.
\]
For a subset of vertices $U \subseteq V$ we define
$w_M(U) \eqdf \sum_{v \in U} w_M(v)$.

@@

\begin{observation}
\label{lm:val-twice}
$w_M(V) = 2 w(M)$.
%\sum_{v \in V} w_M(v) = 2 \sum_{e \in M} w(e)$.
\end{observation}
\begin{proof}
%We have that 
\(\displaystyle
w_M(V)
= \sum_{v \in V} w_M(v)
= \sum_{v \in V} \sum_{(u, v) \in M} w(u, v) +
    \sum_{v \in V} \sum_{(v, u) \in M} w(v, u) 
= 2 \sum_{e \in M} w(e)
%~.    
\).
\end{proof}

Denote by $\delta_M(u,v)$ the difference between the weight of the arc
and the weight of its source vertex, that is:
\[
\delta_M(u, v) \eqdf w(u, v) - w_M(u)
~.
\]
For a subset $S \subseteq A$ of arcs define
$\delta_M(S) \eqdf \sum_{(u,v) \in S} \delta_M(u,v)$.

@@

\begin{definition}
A subset $S_v \subseteq A \cap (V \times \{v\})$ is
an \emph{improvement} with respect to a carpool matching $M$, if
%the following conditions hold:%
\begin{inparaenum}[(i)]
\item $\abs{S_v} \leq c(v)$, and
\item $\delta_M(S_v) > w_M(v)$.
\end{inparaenum}
Furthermore, if there exists an improvement for a vertex $v$, we say
that vertex $v$ can be \emph{improved}.
\end{definition}

Given an arc $(u,v) \in A$, let $\inc(u,v)$ be the set of arcs that
are incident on $(u,v)$, namely define
\[
\inc(u,v)
\eqdf (\nin[](u) \times \{u\}) \cup (\{u\} \times \nout[](u)) \cup
      (\nin[](v) \times \{v\}) \cup (\{v\} \times \nout[](v))
~.
\]
If $S$ is a set of arcs, then $\inc(S) \eqdf \bigcup_{(u,v) \in
S} \inc(u,v)$.
%
Figure~\ref{fig:defs} depicts all the above definitions.

@@

We are now ready to describe our local search algorithm, which is
called \textbf{StarImprove} (Algorithm~\ref{alg:grd}).  It starts with
an empty carpool matching $M$, and in every iteration it looks for a
vertex that can be improved.  If there exists such a vertex $v$, then
the algorithm removes the arcs that are incident on $S_v$ from $M$,
and adds the arcs in $S_v$.  The algorithm terminates when no vertex
can be improved.  Figure~\ref{fig:improvement} depicts an improvement
step.

@@

\end{tikzpicture}
\caption{The matching after improving vertex 3.}
\end{subfigure}
\caption{An improvement example.}
\label{fig:improvement}
\end{figure}

@@

Consider a driver $v \in D_{M^*}$, and let $S = (V \times \{v\}) \cap
M^*$.  By lemma~\ref{lm:no improve} we know that $w(S) \leq w_M(v) +
w_M(\nin[S](v))$, thus we can pay for $S$, using the money on $v$ and
on $\nin[S](v)$.  These vertices are not charged again, since the
stars are disjoint in $M^*$.
\end{proof}

We show that our analysis is tight using Figure~\ref{fig:grd worst}.

\begin{figure}[t]
\centering
\begin{tikzpicture}[every node/.style={default node}, ->]

@@

\end{tikzpicture}
\caption{
%
Consider a directed path with $n$ vertices and $(n-1)$ arcs, where $n$
is even.  The weights for the arcs are alternating between 1 and 2.
%
If \textbf{StarImprove} selects all arcs of weight 1, then no further
improvement can be done and the value of the matching is $n/2$,
while the optimal matching has value of $n-2$.}
\label{fig:grd worst}
\end{figure}

@@

%\section{Group Carpool}
%\label{sec:group}
